Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-03-06 chapters/chapter01.tex
minor changes before it goes to print

% external software
BEAST is packaged with a number of additional utilities that can either assist in generating the input for BEAST, in analyzing the results or in some other specific tasks.
\textbf{BEAUti} (Bayesian Evolutionary Analysis Utility) is a  GUI (graphical user-interface) application for generating XML input files for BEAST. 
% It alleviates some of the most cumbersome tasks that would otherwise have to be done manually, thus saving time and simplifying 
It supports the specification of (standard) models, priors, operators, and MCMC chain settings, without having to resort to a manual editing of the XML syntax.
\textbf{LogCombiner} combines log files from multiple runs, that can then be imported into \textbf{Tracer}, a graphical tool for the analysis and visualization of the output of BEAST as well as other common MCMC packages.
It allows for a quick inspection of mixing, convergence, and summary statistics of the parameters of interest.
\textbf{TreeAnnotator} summarizes the tree log files and generally represents the evolutionary history using a maximum clade credibility (MCC) tree; each node or branch in this tree can be annotated with external information, such as posterior probabilities and maximum a posteriori ancestral traits.
The tree and its annotated information can subsequently be visualized using \textbf{FigTree} (\url{http://tree.bio.ed.ac.uk/software/figtree/}). % or if nodes of the trees are annotated with spatial information they can be displayes in geographical coordinates using \textbf{SPREAD}. %PL we need to introduce spread first, which is a chapter in this thesis :-), same for piBUSS
%Finally \textbf{$\pi$BUSS} is a software for sequence simulation, capable of generating XML files understood by BEAST for combined simulation-inference analysis as well as sequences in formats taht can be imported into any other phylogenetic software.
%
In conclusion, BEAST is a flexible package for statistical pylogenetics and hypothesis testing.
The software is continuously being improved and new models are being developed, but the development has recently been split into a continuing BEAST1 branch and a new BEAST2 branch\citep{Bouckaert2014}. 
Since the work presented in this thesis is implemented in BEAST1, all references to BEAST will implicitly refer to this development branch.
%In addition to a large array of available models, it is actively developed with new models and enhancements included with every new stable release.

@@

\cite{Baele2012} have performed a comparative study on the performance of HME, the stabilized/smoothed HME (sHME), the AICM (a posterior simulation-based analogue of Akaike's information criterion through Markov chain Monte Carlo), PS and SS in a phylogenetic framework.
The authors showed that PS and SS consistently outperformed the harmonic mean estimators as well as the AICM when comparing demographic models as well as relaxed clock models.
Although offering tremendous improvements in bias and variance over the HME both PS and SS require considerable computational effort as the (log) marginal likelihood can no longer be estimated from the likelihood trace that is being output from the standard MCMC analysis. 
Further, great care must also be taken to specify proper priors for all parameters in Equation~\ref{eq:ss_ps}, so as to avoid numerical instabilities when the integration approaches the prior \cite{Baele2013b}.

>>>>

% external software
BEAST is packaged with a number of additional utilities that can either assist in generating the input for BEAST, in analyzing the results or in some other specific tasks.
\textbf{BEAUti} (Bayesian Evolutionary Analysis Utility) is a  GUI (graphical user-interface) application for generating XML input files for BEAST. 
% It alleviates some of the most cumbersome tasks that would otherwise have to be done manually, thus saving time and simplifying 
It supports the specification of (standard) models, priors, operators, and MCMC chain settings, without having to resort to a manual editing of the XML syntax.
\textbf{LogCombiner} combines log files from multiple runs, that can then be imported into \textbf{Tracer}, a graphical tool for the analysis and visualization of the output of BEAST as well as other common MCMC packages.
It allows for a quick inspection of mixing, convergence, and summary statistics of the parameters of interest.
\textbf{TreeAnnotator} summarizes the tree log files and generally represents the evolutionary history using a maximum clade credibility (MCC) tree; each node or branch in this tree can be annotated with external information, such as posterior probabilities and maximum a posteriori ancestral traits.
The tree and its annotated information can subsequently be visualized using \textbf{FigTree} (\url{http://tree.bio.ed.ac.uk/software/figtree/}). % or if nodes of the trees are annotated with spatial information they can be displayes in geographical coordinates using \textbf{SPREAD}. %PL we need to introduce spread first, which is a chapter in this thesis :-), same for piBUSS
%Finally \textbf{$\pi$BUSS} is a software for sequence simulation, capable of generating XML files understood by BEAST for combined simulation-inference analysis as well as sequences in formats taht can be imported into any other phylogenetic software.
%
In conclusion, BEAST is a flexible package for statistical pylogenetics and hypothesis testing.
The software is continuously being improved and new models are being developed, but the development has recently been split into a continuing BEAST1 branch and a new BEAST2 branch \citep{Bouckaert2014}. 
Since the work presented in this thesis is implemented in BEAST1, all references to BEAST will implicitly refer to this development branch.
%In addition to a large array of available models, it is actively developed with new models and enhancements included with every new stable release.

@@

\cite{Baele2012} have performed a comparative study on the performance of HME, the stabilized/smoothed HME (sHME), the AICM (a posterior simulation-based analogue of Akaike's information criterion through Markov chain Monte Carlo), PS and SS in a phylogenetic framework.
The authors showed that PS and SS consistently outperformed the harmonic mean estimators as well as the AICM when comparing demographic models as well as relaxed clock models.
Although offering tremendous improvements in bias and variance over the HME both PS and SS require considerable computational effort as the (log) marginal likelihood can no longer be estimated from the likelihood trace that is being output from the standard MCMC analysis. 
Further, great care must also be taken to specify proper priors for all parameters in Equation~\ref{eq:ss_ps}, so as to avoid numerical instabilities when the integration approaches the prior \citep{Baele2013b}.

### 2015-03-06 chapters/chapter06.tex
minor changes before it goes to print

\paragraph{}
Currently it seems that the semiconductor industry has established two, \textit{nomen set omen}, parallel paths for designing multiprocessors.
The \textit{multicore} path began with the idea of putting two CPU cores on the same chip.
This approach means that the current clock speed of a single core remains the same, but at the same time the vendors are approximately doubling the number of effective cores with every new generation.
In contrast, the \textit{many-core} path focuses on increasing the throughput of parallel applications by using large numbers of much smaller and simpler cores, and again approximately doubling the number of cores that is put on every new generation of devices. 
NVIDIA's GPUs are an exemplar of such many-core devices.
GPUs \citep{Nickolls2008} are making a big impact on the scientific computing, which falls under the term of general-purpose computing on graphics processing units (GPGPU).
%Because modern GPU devices are designed to maximize throughput of a given computation they are capable of achieving impressive peak performances (see Subsection~\ref{sub:fine_grain}).
The most important factor that contributed to the wide adoption of GPUs for scientific computing is the fact that substantial performance improvements are relatively easily affordable. 
Furthermore the parallel computing architecture comes with a variety of libraries, tools and extended C language to leverage the GPGPU capabilities, including NVIDIA's CUDA framework and various implementations of the OpenCL (Open Computing Language) standard.

% beagle
\paragraph{}
%\cite{Suchard2009} have pioneered the use of GPGPU for statistical phylogenetics.
By developing the BEAGLE high-performance library for  evaluating phylogenetic likelihoods of molecular sequence evolution, \cite{Suchard2009} 
have pioneered the use of GPGPU in statistical phylogenetics.
%This work has resulted in the development of the Beagle high-performance library for  evaluating phylogenetic likelihoods of biomolecular sequence evolution \citep{Ayres2012}.
BEAGLE can make use of other highly-parallel processors beside GPUs and is also capable of calling SSE (Streaming SIMD Extensions) as well as AVX (Advanced Vector Extensions) vector instructions, thus achieving improved performance on the current generation of Intel microprocessor architectures.
The BEAGLEAPI \citep{Ayres2012} can currently be used by a range of phylogenetic software tools, such as MrBayes (\url{http://mrbayes.sourceforge.net}), Garli (\url{https://garli.googlecode.com}), PhyML (\url{https://phyml.googlecode.com}) and BEAST.

>>>>

\paragraph{}
Currently it seems that the semiconductor industry has established two, \textit{nomen omen}, parallel paths for designing multiprocessors.
The \textit{multicore} path began with the idea of putting two CPU cores on the same chip.
This approach means that the current clock speed of a single core remains the same, but at the same time the vendors are approximately doubling the number of effective cores with every new generation.
In contrast, the \textit{many-core} path focuses on increasing the throughput of parallel applications by using large numbers of much smaller and simpler cores, and again approximately doubling the number of cores that is put on every new generation of devices. 
NVIDIA's GPUs are an exemplar of such many-core devices.
GPUs \citep{Nickolls2008} are making a big impact on the scientific computing, which falls under the term of general-purpose computing on graphics processing units (GPGPU).
%Because modern GPU devices are designed to maximize throughput of a given computation they are capable of achieving impressive peak performances (see Subsection~\ref{sub:fine_grain}).
The most important factor that contributed to the wide adoption of GPUs for scientific computing is the fact that substantial performance improvements are relatively easily affordable. 
Furthermore the parallel computing architecture comes with a variety of libraries, tools and extended C language to leverage the GPGPU capabilities, including NVIDIA's CUDA framework and various implementations of the OpenCL (Open Computing Language) standard.

% beagle
\paragraph{}
%\cite{Suchard2009} have pioneered the use of GPGPU for statistical phylogenetics.
By developing the BEAGLE high-performance library for  evaluating phylogenetic likelihoods of molecular sequence evolution, \cite{Suchard2009} 
have pioneered the use of GPGPU in statistical phylogenetics.
%This work has resulted in the development of the Beagle high-performance library for  evaluating phylogenetic likelihoods of biomolecular sequence evolution \citep{Ayres2012}.
BEAGLE can make use of other highly-parallel processors beside GPUs and is also capable of calling SSE (Streaming SIMD Extensions) as well as AVX (Advanced Vector Extensions) vector instructions, thus achieving improved performance on the current generation of Intel microprocessor architectures.
The BEAGLE API \citep{Ayres2012} can currently be used by a range of phylogenetic software tools, such as MrBayes (\url{http://mrbayes.sourceforge.net}), Garli (\url{https://garli.googlecode.com}), PhyML (\url{https://phyml.googlecode.com}) and BEAST.
